---
title: "A Layered Grammar of Graphics: ggplot2 methods"
author: "Sun Yiping"
---

## 1. Learning Outcome

In this chapter, we will learn the basic principles and essential components of ggplot2. At the same time, we will gain hands-on experience on using these components to plot statistical graphics based on the principle of Layered Grammar of Graphics.

At the end of this chapter, we will be able to apply the essential graphical elements provided by ggplot2 to create elegant and yet functional statistical graphics.

## 2. Getting Started

### 2.1 Installing and loading the required libraries

Let's first install the required packages for this hands-on exercise.

```{r}
pacman::p_load(tidyverse)
```

### 2.2 Importing the data

The data that would be used for this hands-on exercise is called "Exam_data", which contains the year end examination grades of a cohort of primary 3 students from a local school.

Let's import the data using read_csv() function.

```{r}
exam_data <- read_csv("D:/MITB_SunYP/ISSS608/Data/Exam_data.csv")
```













