---
title: "Visualising Geospatial Point Data"
author: "Sun Yiping"
date: "Febrary 28, 2024"
date-modified: "Febrary 28, 2024"
---

## 1. Learning Outcome

In this hands-on exercise, we will learn how to create a proportional symbol map showing the number of wins by Singapore Pools’ outlets using an R package called *tmap*.

## 2. Getting Started

### 2.1 Installing and loading the required libraries

Firstly, let's install and load the required packages:

-   [tidyverse](https://www.tidyverse.org/): an opinionated collection of R packages designed for data import, data wrangling and data exploration

-   [sf](https://cran.r-project.org/web/packages/sf/index.html): a standardized way to encode spatial vector data.

-   [tmap](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html): makes it easier to plot thematic maps.

```{r}
pacman::p_load(tidyverse, sf, tmap)

```

### 2.2 Importing the data

In this exercise, *SGPools_svy21.csv* file will be used.

Let's start by importing the data.

```{r}
sgpools <- read_csv("../../Data/hands-on_ex07/aspatial/SGPools_svy21.csv")

```

```{r}
list(sgpools) 

```

There are 306 rows and 7 columns in the data file. The coordinates are stored in XCOORD and YCOORD columns, and they follow [Singapore SVY21 Projected Coordinates System](https://app.sla.gov.sg/sirent/About/PlaneCoordinateSystem).

### 2.3 Data Preparation

Next, we'll convert sgpools data frame into a simple feature data frame by using *st_as_sf()* of **sf** package.

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs = 3414)

```

The *crs* argument required us to provide the coordinates system in **EPSG** format. EPSG: 3414 is Singapore SVY21 Projected Coordinate System. We can search for other country’s EPSG code by refering to [epsg.io](https://epsg.io/).

```{r}
list(sgpools_sf)

```

The simple feature data frame now contains 306 rows and 6 columns. The original *XCOORD* and *YCOORD* columns have been converted to *geometry* column.

## 3. Drawing Proportional Symbol Map

Before ploting the map, let's first turn on the interactive mode of *tmap*.

```{r}
tmap_mode("view")

```

### 3.1 It all started with an interactive point symbol map

The code chunks below are used to create an interactive point symbol map.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "red",
           size = 1,
           border.col = "black",
           border.lwd = 1)

```

The red bubbles appeared at the locations where there are data and they are of the same size in the plot above. 

### 3.2 Lets make it proportional

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "red",
           size = "Gp1Gp2 Winnings",
           border.col = "black",
           border.lwd = 1)

```

Now the size of the bubbles vary according to the value of Gp1Gp2 Winnings column.

### 3.3 Lets give it a different colour

We can also assign the color of the bubbles based on another column.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1)

```

Now, the bubbles are colored according to the outlet type. Green for branch and yellow for outlet.

### 3.4 I have a twin brothers :)

We can also put two plots side by side for easy comparison.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)

```

Instead of displaying both outlets on the same plot, we now have one plot for each outlet type. What's more interesting? They are synchronized! If you change the scale of one plot, the other one will also change accordingly. A red circle is also appeared on the other plot pointing to the same area that you are looking at. Isn't this amazing?

Before we end the session, always remember to switch tmap’s Viewer back to plot mode by using the code chunk below.

```{r}
tmap_mode("plot")

```

This comes to the end of this hands-on exercise. I have learned to plot proportional symbol map using in R. Hope you enjoyed it, too!

See you in the next hands-on exercise 🥰






