---
title: "Programming Animated Statistical Graphics with R"
author: "Sun Yiping"
date: "January 26, 2024"
date-modified: "last-modified"
execute: 
  warning: false
---

## 1. Learning Outcome

In this hands-on exercise, we will learn to

-   create animated data visualization by using **gganimate** and **plotly** R packages

-   reshape data using **tidyr** package

-   process, wrangle and transform data using **dplyr** package

## 2. Getting Started

### 2.1 Installing and loading the required libraries

Firstly, let's install and load the required packages:

-   [tidyverse](https://www.tidyverse.org/): an opinionated collection of R packages designed for data import, data wrangling and data exploration

-   [plotly](https://plotly.com/r/): to plot interative statistical graphs

-   [gganimate](https://gganimate.com/): to plot animated visualization

-   [gifski](https://cran.r-project.org/web/packages/gifski/): to convert images to GIF animations

-   [gapminder](https://cran.r-project.org/web/packages/gapminder/): An excerpt of the data available at Gapminder.org. We'll use its *country_colors* scheme.

-   [readxl](https://readxl.tidyverse.org/): to read data from excel files

```{r}
pacman::p_load(tidyverse, plotly, gganimate, gifski, gapminder, readxl)

```

### 2.2 Importing the data

In this hands-on exercise, we'll use the *Data* worksheet from *GlobalPopulation* Excel workbook

Let's start by importing the data.

```{r}
col <- c("Country", "Continent")

globalPop <- read_xls("../../Data/GlobalPopulation.xls",
                      sheet = "Data") %>%
  mutate_each_(funs(factor(.)), col) %>% # to convert all character data into factor
  mutate(Year = as.integer(Year)) # to convert Year column into integer

```

::: callout-important

Note that *mutate_each_()* was deprecated in dplyr 0.7.0. and *funs()* was deprecated in dplyr 0.8.0. Therefore, we will replace the function with *mutate_at()* function. Alternatively, *across()* function can also be used to achieve the same output.

:::

Example code using *mutate_at()* function.

```{r}
col <- c("Country", "Continent")

globalPop <- read_xls("../../Data/GlobalPopulation.xls",
                      sheet = "Data") %>%
  mutate_at(col, as.factor) %>%
  mutate(Year = as.integer(Year))

```

Example code using *across()* function.

```{r}
col <- c("Country", "Continent")

globalPop <- read_xls("../../Data/GlobalPopulation.xls",
                      sheet = "Data") %>%
  mutate(across(col, as.factor)) %>%
  mutate(Year = as.integer(Year))

```

## 3. Animated Data Visualisation: gganimate methods

Now, let's use gganimate methods to create some animated visualizations. The visualizations are generated using the following functions:

-   *transition_()*: defines how the data should be spread out and how it relates to itself across time.

-   *view_()*: defines how the positional scales should change along the animation.

-   *shadow_()*: defines how data from other points in time should be presented in the given point in time.

-   *enter_()*/*exit_()*: defines how new data should appear and how old data should disappear during the course of the animation.

-   *ease_aes()*: defines how different aesthetics should be eased during transitions.

### 3.1 Building a static population bubble plot

Let's first create a static bubble plot using *ggplot2* method.

```{r}
ggplot(globalPop, aes(x = Old, 
                      y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') 

```

### 3.2 Building the animated bubble plot

Next, let's make the bubbles move by using the two functions below:

-   *transition_time()*: to create transition through distinct states in time (i.e. Year).

-   *ease_aes()*: to control easing of aesthetics. The default is linear. Other methods are: *quadratic*, *cubic*, *quartic*, *quintic*, *sine*, *circular*, *exponential*, *elastic*, *back*, and *bounce*.

```{r}
ggplot(globalPop, aes(x = Old, 
                      y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') +
  transition_time(Year) +       
  ease_aes('linear')          

```

## 4. Animated Data Visualisation: plotly

Now, let's make the animated visualization interactive by using *ggplotly()* method.

### 4.1 Building an animated bubble plot: ggplotly() method

::: panel-tabset

## The plot

```{r}
#| echo: false

gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young')

ggplotly(gg)

```

## The code

```{r}
#| eval: false

gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young')

ggplotly(gg)

```

:::

Although we have turned off legend using show.legend = FALSE argument, it stills shows in the plot. To overcome this problem, theme(legend.position='none') should be used as shown in the plot and code chunk below.

::: panel-tabset

## The plot

```{r}
#| echo: false

gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young') + 
  theme(legend.position = 'none')

ggplotly(gg)

```

## The code

```{r}
#| eval: false

gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young') + 
  theme(legend.position = 'none')

ggplotly(gg)

```

:::

### 4.2 Building an animated bubble plot: *plot_ly()* method

Another way to make animated bubble plot is to use *plot_ly()* function.

::: panel-tabset

## The plot

```{r}
#| echo: false

bp <- globalPop %>%
  plot_ly(x = ~Old, 
          y = ~Young, 
          size = ~Population, 
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year, 
          text = ~Country, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers'
          ) %>%
  layout(showlegend = FALSE)

bp

```

## The code

```{r}
#| eval: false

bp <- globalPop %>%
  plot_ly(x = ~Old, 
          y = ~Young, 
          size = ~Population, 
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year, 
          text = ~Country, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers'
          ) %>%
  layout(showlegend = FALSE)

bp

```

:::

This comes to the end of this hands-on exercise. I have learned many different methods to create animated visualizations in R. Hope you enjoyed it, too!

See you in the next hands-on exercise 🥰


