---
title: "Visual Multivariate Analysis with Parallel Coordinates Plot"
author: "Sun Yiping"
date: "Febrary 8, 2024"
date-modified: "Febrary 8, 2024"
execute: 
  warning: false
---

## 1. Learning Outcome

In this hands-on exercise, we will learn how to make parallel coordinates plot.

## 2. Getting Started

### 2.1 Installing and loading the required libraries

Firstly, let's install and load the required packages:

-   [tidyverse](https://www.tidyverse.org/): an opinionated collection of R packages designed for data import, data wrangling and data exploration

-   [GGally](https://cran.r-project.org/web/packages/GGally/index.html): extends 'ggplot2' by adding several functions to reduce the complexity of combining geometric objects with transformed data. 

-   [parallelPlot](https://www.rdocumentation.org/packages/dprep/versions/3.0.2/topics/parallelplot/): Constructs a parallel coordinate plot for a data set with classes in last column.

```{r}
pacman::p_load(tidyverse, GGally, parallelPlot)

```

### 2.2 Importing the data

We'll use the data of [World Happines 2018 report](https://worldhappiness.report/ed/2018/) for this hands-on exercise. The original data is stored in an excel file, and it's been converted to a csv file for easy importing.

Let's start by importing the data.

```{r}
wh  <- read_csv("../../Data/WHData-2018.csv")

```

The data contains 156 rows and 12 columns:

-   2 character variables:
    -   Country
    -   Region
-   10 numerical variables:
    -   Happiness score
    -   Whisker-high
    -   Whisker-low
    -   Dystopia
    -   GDP per capita
    -   Social support
    -   Healthy life expectancy
    -   Freedom to make life choices
    -   Generosity
    -   Perceptions of corruption


## 3. Plotting Static Parallel Coordinates Plot

n this section, we will learn how to plot static parallel coordinates plot by using ggparcoord() of GGally package. 

### 3.1 Plotting a simple parallel coordinates

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))

```

It's quite obvious to see that there are some outliers in the following columns:

-   Social support
-   Freedom to make life choices
-   Generosity
-   Percveptions of corruption

### 3.2 Plotting a parallel coordinates with boxplot

However, it's quite difficult understand the distribution of the columns. We can overcome this by adding boxplots.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")

```

We can now make more statistical inference from the plot.

### 3.3 Parallel coordinates with facet

We can further break it down by facet to visualize the differences among different categories.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)

```

Now we have made the parallel coordinates plots for each region.

### 3.4 Rotating x-axis text label

Let's now customize the plots a bit to make the x-axis labels easier to be read.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))

```
```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))

```

## 4. Plotting Interactive Parallel Coordinates Plot: parallelPlot methods

In this section, we will learn how to use functions provided in parallelPlot package to build interactive parallel coordinates plot.

### 4.1 The basic plot

The code chunk below plot an interactive parallel coordinates plot by using parallelPlot().

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)

```

When we click on any data point, the data from the same row is high lighted.

### 4.2 Changing the colour scheme

Let's change the looks of the plot such as rotating the x-axis labels:

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)

```

### 4.3 Parallel coordinates plot with histogram

Similarly, we can add histograms on the chart to view the distributions.

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)

```

This comes to the end of this hands-on exercise. I have learned to make static and interactive parallel coordicates plots in R. Hope you enjoyed it, too!

See you in the next hands-on exercise 🥰  


















